Liveness-Based Garbage Collection for Lazy Languages

### 2016-02-03 paper/paper.tex
rewrite

\todelete{While this idea has been shown  to be effective for a first-order {\em
  eager}  language~\cite{asati14lgc}, a  straightforward extension  of
the technique is not possible for lazy languages, where heap-allocated
objects may  include closures (runtime representations  of unevaluated
expressions).}
We encounter significant challenges in defining a liveness based garbage 
collector for lazy languages. 
\added{
Firstly, during garbage collection the heap contains both fully 
evaluated data and unevaluated thunks/closures. 
For liveness-based
garbage collection to  be effective, we need to extend  it to closures
apart from  evaluated data. Since closures can escape the scope in which 
they were created, they need to carry liveness information of its free 
variables. 
Moreover, since  data  is made  live  by evaluation  of
closures, and  in lazy languages the  place in the program  where this
evaluation  takes  place  cannot be  statically  determined,  laziness
complicates  liveness analysis  itself.  
Also, as  execution progresses and possible  future uses are
eliminated, we  update the  liveness information in  a closure  with a
more precise version.  Because of these reasons, the garbage collector
also  becomes significantly  more  complicated  than a  liveness-based
collector for a eager language.
}

\todelete{
Firstly, since  data  is made  live  by evaluation  of
closures, and  in lazy languages the  place in the program  where this
evaluation  takes  place  cannot be  statically  determined,  laziness
complicates  liveness analysis  itself.  Moreover,  for liveness-based
garbage collection to  be effective, we need to extend  it to closures
apart from  evaluated data.  Since a  closure can escape the  scope in
which it was created, it has  to carry the liveness information of its
free variables. As  execution progresses and possible  future uses are
eliminated, we  update the  liveness information in  a closure  with a
more precise version.  Because of these reasons, the garbage collector
also  becomes significantly  more  complicated  than a  liveness-based
collector for a eager language.
}
Experiments    with   a    single    generation   copying    collector
(Section~\ref{sec:experiments})   confirm  the   expected  performance
benefits.  Liveness-based collection results in an increase in garbage
reclaimed.  As  a consequence, there is  a reduction in the  number of
collections from 1.6X  to 23X and a decrease  in the minimum
memory requirement from  1X to 1389X.  As  an added benefit,
there is also a reduction in the overall execution time in 5 out of 12
benchmark programs.

>>>>

\todelete{While  this  idea has  been  shown  to  be effective  for  a
  first-order     {\em     eager}    language~\cite{asati14lgc},     a
  straightforward extension of the technique  is not possible for lazy
  languages,  where   heap-allocated  objects  may   include  closures
  (runtime   representations    of   unevaluated    expressions).    }
\added{Designing a liveness based garbage collector for lazy languages
  poses significant challenges.}  Firstly, since  data is made live by
evaluation of closures, and in lazy languages the place in the program
where  this evaluation  takes place  cannot be  statically determined,
laziness   complicates  liveness   analysis  itself.    Moreover,  for
liveness-based garbage collection  to be effective, we  need to extend
it to closures apart from evaluated  data.  Since a closure can escape
the  scope in  which it  was  created, it  has to  carry the  liveness
information  of  its  free  variables.  As  execution  progresses  and
possible  future   uses  are   eliminated,  we  update   the  liveness
information  in a  closure with  a more  precise version.   Because of
these reasons,  the garbage collector also  becomes significantly more
complicated than a liveness-based collector for a eager language.

Experiments    with   a    single    generation   copying    collector
(Section~\ref{sec:experiments})   confirm  the   expected  performance
benefits.  Liveness-based collection results in an increase in garbage
reclaimed.  As  a consequence, there is  a reduction in the  number of
collections from 1.6X  to 23X and a decrease  in the minimum
memory requirement from  1X to 1389X.  As  an added benefit,
there is also a reduction in the overall execution time in 5 out of 12
benchmark programs.
